---
title: "Markov state models of a 1D trajectory"
author: "toni.giorgino@cnr.it - Consiglio Nazionale delle Ricerche"
date: "`r date()`"
output:
  html_document:
    number_sections: yes
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(lattice)
```



# Markov modeling on a large dataset

This tutorial shows the Markov model-based computation "macroscopic" quantities 
in a realistic system, simulated with extensive MD. 


There are 3 data sets already downloaded for you:
* `/mnt/scratch/shared/markov/binding` Ligand binding
* `/mnt/scratch/shared/markov/villin`  Villin folding
* `/mnt/scratch/shared/markov/cxcl12`  GPCR conformational transition


We now turn to the files contained in `/mnt/scratch/shared/markov/binding/1/filtered/` (Thrombin
binding dataset). This is 1/3 of the full dataset (see directories 2 and 3).



## Dataset layout

Each `.xtc` file is one trajectory of 20 ns (200 frames). Frames are spaced 100 ps from each other. Each trajectory is a small piece of the whole simulation set.  File names indicate the identity of the trajectory (e.g. `e10s1`) and where it started from (e.g. `e7s5f133` means that the first frame of that configuration was frame 133 of `e7s5`). Water molecules were removed ("filtered").

```
/mnt/scratch/shared/markov/binding/
    ├── 1
    │   └── filtered
    │       ├── e10s1_e7s5f133
    │       │   └── e10s1_e7s5f133-SDOERR_thrombinLig6x1-0-1-RND6286_9.filtered.xtc
    │       ├── e10s2_e7s5f159
    │       │   └── e10s2_e7s5f159-SDOERR_thrombinLig6x1-0-1-RND8907_9.filtered.xtc
    │       ├── e10s3_e7s5f133
    │       │   └── e10s3_e7s5f133-SDOERR_thrombinLig6x1-0-1-RND5451_9.filtered.xtc
...    

            ├── e9s6_e2s5f112
            │   └── e9s6_e2s5f112-SDOERR_thrombinLig6x3-0-1-RND3971_9.filtered.xtc
            ├── filtered.pdb
            └── filtered.psf

860 directories, 858 files
```

## Metric calculation

In the previous steps we used PLUMED to compute the three-dimensional position of the inhibitor
relative to the target. The distance was computed **after** alignment 
with a reference structure (to avoid rotational diffusion).
The trajectories are identified by the first column. 

```{r}
d<-read.table("metrics.dat",head=F)
names(d) <- c("TrajLong","Frame","X","Y","Z")
d$Traj <- sub("_.+$","",d$TrajLong)  # Shorten the trajectory name
d$Traj <- factor(d$Traj)
d$TrajLong <- NULL

head(d)
```

We plot a few of them. Extract the first trajectory for visualization.

```{r}
dp <- subset(d,d$Traj == "e36s5")
matplot(dp$Frame, dp[,c("X","Y","Z")], type="l")

```

...and the second...

```{r}

dp <- subset(d,d$Traj == "e12s5")
matplot(dp$Frame, dp[,c("X","Y","Z")], type="l")


```


## Clustering

Before we proceed, we need to move from the N-dimensional space of continuous
coordinates (in this case 3D) to a discrete "microstate". This is
achieved by clustering.  Of several algorithms, we use k-means.
The Markov model will approximate
the "jumping" process between those states.

```{r}
K <- 100
ustates <- kmeans(d[,c("X","Y","Z")],   K)
d$Microstate <- factor(ustates$cluster)
lab <- 1:K
```


This is how microstates distribution looks in the XY plane.

```{r}
plot(d$X,d$Y,cex=.1,col=d$Microstate)
```

Also in 3D (don't worry if it doesn't work)

```{r}
#install.packages("rgl")
library(rgl)
plot3d(d$X,d$Y,d$Z,
       aspect=T,col=d$Microstate)
```



We can gather statistics on the microstates
```{r}
h<-table(d$Microstate)

plot(h, xlab="Microstate", ylab = "Count")

```

Boltzmann inversion (note that $k_B T \simeq 0.6$ kcal/mol) returns
the $\Delta G$ (up to an irrelevant additive constant).
Too many microstates to have a clear picture.

```{r}
plot(5-.6*log(h),
     xlab="State (X coordinate, 1 to 100)",
     ylab="ΔG (kcal/mol) = -log(count)"
     )

```

Let's look at the most visited microstate.

```{r}
max(h)
which.max(h)

dmost <- subset(d,Microstate==which.max(h))

tmp<-table(dmost$Traj)
tmp[tmp>0]

```


# Markov state modeling

## Construction of the transition matrix

We compute how many times each state to state ($i\to j$)
transition has occurred *after a lag time of $\tau$*.
We need to treat the trajectories separately (tapply).

```{r}
tau <- 10

TT.list <-
tapply(d$Microstate, d$Traj, function(x) {
            table( head(x,-tau),
                   tail(x,-tau),exclude=NULL) 
    } )
```

Then aggregate them by summing the independent transition count matrix, one
per trajectory.

```{r}
TT <- TT.list[[1]]*0
for (m in TT.list) {
    TT <- TT+m
}

```


The transition probability matrix is large (KxK)
but we may visualize it as an image.

```{r}
image(lab,lab,TT,
      xlab="Source state",ylab="Destination state")
```

Of course, transitions around the two
free energy wells are sampled more often, so they show
up clearly.

We divide each row of the transition
count  matrix so that they
sum to unity. We thus obtain the
transition *probability* matrix. Note
that, for simplicity, this is a simplified
treatment in the following respects: (1) 
one should take care
of reversibility, (2) use a maximum-likelihood
estimator, and (3) deal properly with zero counts
(which cause singularities).

Here I am adding a small *pseudocount* to all
counts to avoid null cells.


```{r}

pseudocount <- 0.01
sum1 <- function(x) (x/sum(x))
P <- t(apply(TT+pseudocount,2,sum1 ))
image(lab,lab,P,
      xlab="Source state",ylab="Destination state")

```



## Eigenvalues

From the transition probability matrix **at lag $\tau$**
we compute the *left*  eigenvectors and eigenvalues.

The eigenvalues correspond to the relaxation 
rates towards equilibrium. The largest eigenvalue
is unity, corresponding to the stationary state.

Eigenvalues describe the negative *exponential* relaxation time scales, in units of $\tau$. Therefore, 
the should be converted to "normal" time units
throught the *implied time scales* (ITS) equation, i.e.:

$$\tau_i = -\tau/\log(\mu_i)$$

where $\tau_i$ is the $i$-th implied time scale and $\mu_i$
the corresponding eigenvalue.


```{r fig.height=4}
ei <- eigen(t(P))
eva <- sort(abs(ei$values),decreasing = T)
eve <- ei$vectors

taus <- -tau/log(eva[-1])

plot(eva,ylab="Eigenvalue magnitue",xlab="Eigenvalue index (sorted)")

plot(c(NA,taus),ylab="Implied timescale")
abline(h=tau,col="red")

```

Given the  spread of timescales is usually
large, we plot them in logarithmic scale.

Kinetics faster than $\tau$ can **not** be 
resolved by construction of the transition
matrix.

```{r}

plot(c(NA,taus),ylab="Implied timescale (log)",log="y")
abline(h=tau,col="red")
text(50,5,"Kinetics faster than τ\nnot resolved by construction",
     col="red")

```


## Eigenvectors


Eigenvectors matching a given eigenvalue describe 
the "contribution" of each state to that relaxation
mode.  The eigenvector of the unitary eigenvalue
is the stationary distribution, i.e. the equilibrium
(after normalizing it so it sums to 1).
Its Boltzmann inversion provides the free energy landscape.



```{r}

ev0 <- abs(eve[,1])
ev0dg <- -log(ev0)+max(log(ev0))-2

plot(lab,sum1(ev0),
     xlab="State",ylab="Equilibrium probability")

```


Where shall we find the most probable state? 

```{r}

subset(d,Microstate==which.max(ev0))

```


The other eigenvector describe the 
"wave" of the relaxation (at the matching
implied timescale).

```{r}

ev1 <- 2*scale(Re(eve[,2]))
plot(lab,ev1,
     xlab="State",ylab="Contribution (au)",
     main="Slowest relaxation mode")
polygon(c(0,lab,K+1),c(0,ev1,0),col="gray")
#lines(lab,sum1(ev0)*20,col="red",lty=2)
lines(lab,ev0dg,col="red",lty=2)
mtext(sprintf("%.2f time units",taus[1]),3)
legend("bottomleft",col=1:2,lty=1:2,
       c("Second eigenvector",
         "Free energy profile"))

ev2 <- 2*scale(Re(eve[,3]))
plot(lab,ev2,
     xlab="State",ylab="Contribution (au)",
     main="2nd slowest mode")
polygon(c(0,lab,K+1),c(0,ev2,0),col="gray")
lines(lab,ev0dg,col="red",lty=2)
legend("bottomleft",col=1:2,lty=1:2,
       c("Third eigenvector",
         "Free energy profile"))
mtext(sprintf("%.2f time units",taus[2]),3)

```

The shape of  the eigenvectors suggests
a clustering into coarser-grained macrostates.
This constitutes the gist of 
the Perron-cluster analysis (PCCA+).



## Choice of the lag time

Remember that the transition matrix 
was, by construction,    **a function of $\tau$**.
Therefore, the whole eigenvector analysis 
shown before should be repeated at increasing
values of $\tau$ to study its sensitivity.

We pick the smallest $\tau$ such that the implied
timescales are effectively "flat". Again,
the literature provides much more sophisticated
statistical tests
on the Markovianity assumption (e.g. Chapman-Kolmogorov).
See e.g. [PyEmma's Model selection and validation](http://www.emma-project.org/latest/generated/model_selection_validation.html) tutorial.



```{r}

evanalysis <- function(tau) {
    
    TT.list <-
    tapply(d$Microstate, d$Traj, function(x) {
            table( head(x,-tau),
                   tail(x,-tau),exclude=NULL) 
    } )
    TT <- TT.list[[1]]*0
    for (m in TT.list) {
        TT <- TT+m
    }
    TT <- TT+pseudocount

    P <- t(apply(TT,2,sum1 ))
    ei <- eigen(t(P))
    eva <- sort(abs(ei$values),decreasing = T)
    eve <- ei$vectors
    taus <- -tau/log(eva[-1])
    list(eva=eva,eve=eve,taus=taus)
}

lag_list <- c(1,2,5,10,20,50,100)

ntaus <- 6 # Consider this many eigenvectors
tau_matrix <- matrix(0,length(lag_list),ntaus)
li <- 1
for(lag in lag_list) {
    etaus <- evanalysis(lag)$taus
    tau_matrix[li,] <- head(etaus,ntaus)
    li <- li+1
}

matplot(lag_list,tau_matrix,type="o",log="xy",
        xlab="Lag time τ",ylab="Implied timescales at τ")
lines(1:100,1:100,lwd=4,col="grey",lty=2)

```


So in this case the microstate discretization is not good enough.
We need to try one of those

 * Better starting variables (e.g. include orientation)
 * Use of TICA
 * Perform an higher level (macro-state) clustering
 
All of this complex to implement from scratch. It is better
to use specialized packages such as the ones we'll see in
the next part.





